The Unicode Cookbook \linebreak for Linguists

### 2016-01-07 book/chapters/orthography_profiles.tex
adding R implementation info

\section{Examples}

@@

<<>>==
test <- c('AABB', '\u0041\u0410\u0042\u0412')
write.profile(test)
@

@@

<<>>==
test <- c('AABB', '\u0041\u0410\u0042\u0412')
tokenize( test, profile = c('A', 'B') )
@

>>>>

To illustrate the practical application of orthography profiles, we have
implemented two different versions of the above specifications, one in Python
and one in R. These two implementations have rather different implementation
histories, and we are not 100\% sure that they will in all situations give the
same results. Also the perfomance with larger datasets differs, and the code is
not always as clean as we would like it to be. In sum, the two implementations
should be considered as `proof of concept' and not as the final word on the
practical application of the specifications above. In our own experience, the
current implementations are sufficiently fast and stable to be useful for
academic practice (e.g. checking data consistency, or analyzing and
transliterating small to medium sized data sets), but they should probably
better not be used for full-scale industry applications.

\subsection*{Installing the R implementation}

The R implementation is available in the package \texttt{qlcData}, which is 
directly available from the central R repository CRAN (Comprehensive R Archive 
Network). The R software environment itself has to be downloaded from its 
website.\footnote{\url{https://www.r-project.org}} After starting the included 
R program, the \texttt{qlcData} package for dealing with orthography profiles can be 
simply installed as follows:

<<eval=FALSE>>==
# download and install the qlcData software
install.packages('qlcData') 
# load the software, so it can be used
library(qlcData) 
@

\noindent The version available through CRAN is the most recent stable version.
To obtain the most recent bug-fixes and experimental additions, please use the
development versions, which is available on
GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
version can be easily installed using the github-install helper software from the
\texttt{devtools} package.

<<eval=FALSE>>==
# download and install helper software
install.packages('devtools') 
# install the qlcData package from GitHub
devtools::install_github('cysouw/qlcData')
# load the software, so it can be used 
library(qlcData) 
@

\noindent Inside the \texttt{qlcData} package, there are two functions for
orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
help files with illustrative examples, and also a so-called `vignette' with
explanations and examples.

<<eval=FALSE>>==
# view help files
help(write.profile)
help(tokenize)
# view vignette with explanation and examples
vignette('orthography_processing')
@

\noindent Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed inside R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) `shiny' webapps.

The bash executables are little files providing an interface to the R code that
can be used in a shell on a UNIX-alike machine. The location of these
executables is a bit hidden away by the install procedure of R packages. The
location can be found by the following command in R. These executables can be 
used from here, or they can be moved and/or copied to any location as wanted.

<<eval=FALSE>>==
# show the path to the bash executables
file.path(find.package('qlcData'), 'exec')
@

\noindent Another option is to add an alias for these executables to the bash
profile. The following R function can help adding an alias.

<<eval=FALSE>>==
# help function to add an alias to the bash profile
add_alias('tokenize')
@

After this addition
it should be possible to access the \texttt{tokenize} function from the shell.
Try \texttt{tokenize --help} to test the functionality.

<<help, engine='bash'>>=
tokenize --help
@

To make the functionality even more accessible, we have prepared webapps with 
the \texttt{shiny} framework for the R functions. These webapps are available 
online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
package and can be started with the following helper function:

<<eval=FALSE>>==
launch_shiny('tokenize')
@

\subsection*{Installung the Python implementation}

TO DO

\section{Abstract examples}

\subsection*{Error reporting}

@@

<<>>=
test <- c('AABB', '\u0041\u0410\u0042\u0412')
write.profile(test)
@

@@

<<>>=
test <- c('AABB', '\u0041\u0410\u0042\u0412')
tokenize( test, profile = c('A', 'B') )
@

\subsection*{Different ways to write a profile}

The preparation of an orthography profile from some data might sound like a 
trivial problem, but actually there are various different ways in which strings 
can be separated into graphemes by \texttt{write.profile}. Consider the 
following example string. The default settings of \texttt{write.profile} 
separates the string into Unicode graphemes according to grapheme clusters 
(`user-perceived characters', see Section~\ref{the-unicode-approach}).

<<>>=
example <- '\u00d9\u00da\u00db\u0055\u0300\u0055\u0301\u0055\u0302'
profile_1 <- write.profile(example)
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_1) <- c('Gr.', 'Freq.', 'Codepoint', 'Unicode Name')
print(xtable(profile_1
        , caption = 'Profile 1 (default settings, splitting grapheme clusters)'
        , label = 'tab:profile1'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

\noindent Some of these graphemes are single codepoints, others are combinations
of two codepoints. By specifying the separator as empty, it is possible to split
the string into codepoints.

<<>>=
profile_2 <- write.profile(example, sep = "")
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_2) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
print(xtable(profile_2
        , caption = 'Profile 2 (splitting by codepoints)'
        , label = 'tab:profile2'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

\noindent Some characters look identical, although they are encoded differently. 
Unicode offers different ways of normalization, which can be invoked here as 
well (see Section~\ref{pitfall-canonical-equivalence}).

<<>>=
# after NFC normalization unicode codepoints have changed
profile_3 <- write.profile(example, normalize = "NFC", sep = "")
# NFD normalization gives yet another structure of the codepoints
profile_4 <- write.profile(example, normalize = "NFD", sep = "")
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_3) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
print(xtable(profile_3
        , caption = 'Profile 3 (splitting by NFC codepoints)'
        , label = 'tab:profile3'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_4) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
print(xtable(profile_4
        , caption = 'Profile 4 (splitting by NFD codepoints)'
        , label = 'tab:profile4'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

\noindent It is important to realize that for Unicode grapheme definitions, NFC 
and NFD normalization are equivalent. This can be shown by normalizing the 
example in either NFC or NFD, but using default separation in 
\texttt{write.profile}. To be precise, default separation means setting \texttt{sep = NULL} 
but that has not be added explicitly.

<<>>=
# note that NFC and NFD normalization are identical
# for unicode grapheme definitions
profile_5 <- write.profile(example, normalize = "NFD")
profile_6 <- write.profile(example, normalize = "NFC")
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_5) <- c('Gr.', 'Freq.', 'Codepoint', 'Unicode Name')
print(xtable(profile_5
        , caption = 'Profile 5 (splitting by graphemes after NFD)'
        , label = 'tab:profile5'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_6) <- c('Gr.', 'Freq.', 'Codepoint', 'Unicode Name')
print(xtable(profile_6
        , caption = 'Profile 6 (splitting by graphemes after NFC)'
        , label = 'tab:profile6'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

### 2016-01-07 book/unicode.tex
adding R implementation info

\include{chapters/introduction}
\include{chapters/pitfalls}
\include{chapters/ipa_background}
\include{chapters/ipa_meets_unicode}

@@

\section{Examples}

@@

@@

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Thu Dec 10 18:07:34 2015
\begin{table}[H]
\centering
{\small
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & UnicodeName \\ 
  \midrule
A & 1 & U+0041 & LATIN CAPITAL LETTER A \\ 
  Α & 1 & U+0391 & GREEK CAPITAL LETTER ALPHA \\ 
  А & 1 & U+0410 & CYRILLIC CAPITAL LETTER A \\ 
   \bottomrule
\end{tabular}
}
\caption{test} 
\label{bla}
\end{table}

>>>>

\include{chapters/preface}
\include{chapters/introduction}
\include{chapters/pitfalls}
\include{chapters/ipa_background}
\include{chapters/ipa_meets_unicode}

@@

To illustrate the practical application of orthography profiles, we have
implemented two different versions of the above specifications, one in Python
and one in R. These two implementations have rather different implementation
histories, and we are not 100\% sure that they will in all situations give the
same results. Also the perfomance with larger datasets differs, and the code is
not always as clean as we would like it to be. In sum, the two implementations
should be considered as `proof of concept' and not as the final word on the
practical application of the specifications above. In our own experience, the
current implementations are sufficiently fast and stable to be useful for
academic practice (e.g. checking data consistency, or analyzing and
transliterating small to medium sized data sets), but they should probably
better not be used for full-scale industry applications.

\subsection*{Installing the R implementation}

The R implementation is available in the package \texttt{qlcData}, which is 
directly available from the central R repository CRAN (Comprehensive R Archive 
Network). The R software environment itself has to be downloaded from its 
website.\footnote{\url{https://www.r-project.org}} After starting the included 
R program, the \texttt{qlcData} package for dealing with orthography profiles can be 
simply installed as follows:

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# download and install the qlcData software}
\hlkwd{install.packages}\hlstd{(}\hlstr{"qlcData"}\hlstd{)}
\hlcom{# load the software, so it can be used}
\hlkwd{library}\hlstd{(qlcData)}
\end{alltt}
\end{kframe}
\end{knitrout}

\noindent The version available through CRAN is the most recent stable version.
To obtain the most recent bug-fixes and experimental additions, please use the
development versions, which is available on
GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
version can be easily installed using the github-install helper software from the
\texttt{devtools} package.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# download and install helper software}
\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
\hlcom{# install the qlcData package from GitHub}
\hlstd{devtools}\hlopt{::}\hlkwd{install_github}\hlstd{(}\hlstr{"cysouw/qlcData"}\hlstd{)}
\hlcom{# load the software, so it can be used}
\hlkwd{library}\hlstd{(qlcData)}
\end{alltt}
\end{kframe}
\end{knitrout}

\noindent Inside the \texttt{qlcData} package, there are two functions for
orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
help files with illustrative examples, and also a so-called `vignette' with
explanations and examples.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# view help files}
\hlkwd{help}\hlstd{(write.profile)}
\hlkwd{help}\hlstd{(tokenize)}
\hlcom{# view vignette with explanation and examples}
\hlkwd{vignette}\hlstd{(}\hlstr{"orthography_processing"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\noindent Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed inside R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) `shiny' webapps.

The bash executables are little files providing an interface to the R code that
can be used in a shell on a UNIX-alike machine. The location of these
executables is a bit hidden away by the install procedure of R packages. The
location can be found by the following command in R. These executables can be 
used from here, or they can be moved and/or copied to any location as wanted.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# show the path to the bash executables}
\hlkwd{file.path}\hlstd{(}\hlkwd{find.package}\hlstd{(}\hlstr{"qlcData"}\hlstd{),} \hlstr{"exec"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\noindent Another option is to add an alias for these executables to the bash
profile. The following R function can help adding an alias.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# help function to add an alias to the bash profile}
\hlkwd{add_alias}\hlstd{(}\hlstr{"tokenize"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

After this addition
it should be possible to access the \texttt{tokenize} function from the shell.
Try \texttt{tokenize --help} to test the functionality.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
tokenize --help
\end{alltt}

\begin{verbatim}
## Error in running command bash
\end{verbatim}
\end{kframe}
\end{knitrout}

To make the functionality even more accessible, we have prepared webapps with 
the \texttt{shiny} framework for the R functions. These webapps are available 
online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
package and can be started with the following helper function:

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{launch_shiny}\hlstd{(}\hlstr{"tokenize"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\subsection*{Installung the Python implementation}

TO DO

\section{Abstract examples}

\subsection*{Error reporting}

@@

\subsection*{Different ways to write a profile}

The preparation of an orthography profile from some data might sound like a 
trivial problem, but actually there are various different ways in which strings 
can be separated into graphemes by \texttt{write.profile}. Consider the 
following example string. The default settings of \texttt{write.profile} 
separates the string into Unicode graphemes according to grapheme clusters 
(`user-perceived characters', see Section~\ref{the-unicode-approach}).

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{example} \hlkwb{<-} \hlstr{"ÙÚÛÙÚÛ"}
\hlstd{profile_1} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)}
\end{alltt}
\end{kframe}
\end{knitrout}

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Thu Jan  7 01:22:44 2016
\begin{table}[H]
\centering
{\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ú & 1 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Ù & 1 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
  Û & 1 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
}
\caption{Profile 1 (default settings, splitting grapheme clusters)} 
\label{tab:profile1}
\end{table}

\noindent Some of these graphemes are single codepoints, others are combinations
of two codepoints. By specifying the separator as empty, it is possible to split
the string into codepoints.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{profile_2} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Thu Jan  7 01:22:44 2016
\begin{table}[H]
\centering
{\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 1 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 1 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 1 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 3 & U+0055 & LATIN CAPITAL LETTER U \\ 
  Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
}
\caption{Profile 2 (splitting by codepoints)} 
\label{tab:profile2}
\end{table}

\noindent Some characters look identical, although they are encoded differently. 
Unicode offers different ways of normalization, which can be invoked here as 
well (see Section~\ref{pitfall-canonical-equivalence}).

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# after NFC normalization unicode codepoints have changed}
\hlstd{profile_3} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFC"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
\hlcom{# NFD normalization gives yet another structure of the codepoints}
\hlstd{profile_4} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Thu Jan  7 01:22:44 2016
\begin{table}[H]
\centering
{\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
}
\caption{Profile 3 (splitting by NFC codepoints)} 
\label{tab:profile3}
\end{table}

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Thu Jan  7 01:22:44 2016
\begin{table}[H]
\centering
{\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 2 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 2 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 2 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 6 & U+0055 & LATIN CAPITAL LETTER U \\ 
   \bottomrule
\end{tabular}
}
\caption{Profile 4 (splitting by NFD codepoints)} 
\label{tab:profile4}
\end{table}

\noindent It is important to realize that for Unicode grapheme definitions, NFC 
and NFD normalization are equivalent. This can be shown by normalizing the 
example in either NFC or NFD, but using default separation in 
\texttt{write.profile}. To be precise, default separation means setting \texttt{sep = NULL} 
but that has not be added explicitly.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# note that NFC and NFD normalization are identical for unicode grapheme}
\hlcom{# definitions}
\hlstd{profile_5} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{)}
\hlstd{profile_6} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFC"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Thu Jan  7 01:22:44 2016
\begin{table}[H]
\centering
{\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 2 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 2 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
}
\caption{Profile 5 (splitting by graphemes after NFD)} 
\label{tab:profile5}
\end{table}

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Thu Jan  7 01:22:44 2016
\begin{table}[H]
\centering
{\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
}
\caption{Profile 6 (splitting by graphemes after NFC)} 
\label{tab:profile6}
\end{table}

@@

% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Wed Jan  6 22:54:44 2016
\begin{table}[H]
\centering
{\small
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & UnicodeName \\ 
  \midrule
A & 1 & U+0041 & LATIN CAPITAL LETTER A \\ 
  Α & 1 & U+0391 & GREEK CAPITAL LETTER ALPHA \\ 
  А & 1 & U+0410 & CYRILLIC CAPITAL LETTER A \\ 
   \bottomrule
\end{tabular}
}
\caption{test} 
\label{bla}
\end{table}
